Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-18 mr-review.Rmd
organizing text...

Many approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. Both two-stage strategies and likelihood-based strategies 
make parametric assumptions for the identification of causal effects [@burgess_review_2017].

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).

@@

## Assumption violations

@@

## Sensitivity analysis for MR with one instrument and one outcome

@@

probabilistic Mendelian randomization (PMR-Egger) for two-sample MR with correlated SNPs models [@yuan_testing_2020]

Early MR methods account manage horizontal pleiotropy by removing instrumental SNPs 
that may be associated with the outcome variable [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].

@@

CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
affect performance. 
CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
models as an effect on a shared, unobserved factor. The remaining SNPs are independent of this 
unobserved factor. Every SNP can have a nonzero pleiotropic effect on the outcome, and these 
pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
to determine whether the GWAS summary statistics are consistent with a causal effect of the        
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
causal effect [@vehtari_practical_2017].

MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref() and \@ref()).

>>>>

Many approaches to MR can be classified into two categories: two-stage methods and 
likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
outcome, one may perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).

@@

# Assumption violations

@@

# Sensitivity analysis

@@

@@

CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021]. 
Their approaches to modeling correlated horizontal pleiotropy differ in important ways that ultimately 
affect performance. 
CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
models as an effect on a shared, unobserved factor. The remaining SNPs are independent of this 
unobserved factor. Every SNP can have a nonzero pleiotropic effect on the outcome, and these 
pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
to determine whether the GWAS summary statistics are consistent with a causal effect of the        
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
causal effect [@vehtari_practical_2017].

MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).
